---
title: "Illusions: Hierarchical Modeling Tutorial"
author: "Mahbod Mehrvarz"
date: "August 2025"
output:
  pdf_document:
    latex_engine: pdflatex
    toc_depth: 2
fontsize: 12pt
geometry: margin=1in
header-includes:
  - \usepackage{amsmath}\numberwithin{equation}{section}
  - \usepackage{hyperref}\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}

---


This note accompanies *Localizing Structure in Individual Differences: A Visual Illusion Case Study* (Mehrvarz, Popat, Rouder). It shows how to fit two hierarchical models to the illusions data: a Gaussian baseline with an inverse-Wishart prior and a hierarchical Bayesian factor model.

# Overview

Goal: fit models that separate trial noise from across-individual structure.

1. load and standardize data
2. Gaussian baseline with IW prior (population vs individual correlations)
3. hierarchical factor model with simple identification
4. posterior summaries, tables, and plots

**Packages:** `R2jags`, `abind`, `infinitefactor`, `rlist`, `RColorBrewer`.

```{r setup, message=FALSE, warning=FALSE}
set.seed(42)
options(width = 90)
need = c("R2jags","abind","infinitefactor","rlist","RColorBrewer","corrplot")
to_install = setdiff(need, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install, quiet = TRUE)
library(R2jags); library(abind); library(infinitefactor); library(rlist) 
library(RColorBrewer)
```

# Data

```{r}
url1 = "https://raw.githubusercontent.com/specl/ind-illusions/"
url2 = "refs/heads/public/_data/clean_data.csv"
url = paste0(url1, url2)
dat = read.csv(url)

dat$S  = as.numeric(factor(dat$sub))
dat$J0 = as.numeric(factor(dat$task))      # 1..5 tasks
dat$JV = (dat$J0 - 1) * 2 + dat$version    # 1..10 task×version

# variance-normalize by task×version
v1 = tapply(dat$y, list(dat$S, dat$JV), var)
divisor = sqrt(apply(v1, 2, mean))
dat$yc = dat$y / divisor[dat$JV]

I = length(unique(dat$S))
J = 10                                     # task×version
D = 7                                      # 5 task factors + 2 general
N = nrow(dat)

task_names = c("Brentano 1","Brentano 2","Ebbinghaus 1","Ebbinghaus 2",
               "Poggendorf 1","Poggendorf 2","Ponzo 1","Ponzo 2",
               "Zoellner 1","Zoellner 2")
```

---

# Gaussian baseline with inverse-Wishart prior (before factors)

We fit a hierarchical Gaussian that captures two correlation sources.

For $y\_{ijk}$ with person $i$, indicator $j$ (task×version), replicate $k$:

$$
\begin{aligned}
y_{ijk}\mid \theta_{ij},\,\tau_j^2 &\sim \mathcal{N}\!\big(\theta_{ij},\,\tau_j^2\big),\\
\boldsymbol{\theta}_i \mid \boldsymbol{\mu},\,\boldsymbol{\Omega}
&\sim \mathcal{N}_J\!\big(\boldsymbol{\mu},\,\boldsymbol{\Omega}^{-1}\big).
\end{aligned}
$$

Priors:

$$
\mu_j \sim \mathcal{N}(\mu_m,\mu_s^2),\quad
\tau_j^{-2} \sim \mathrm{Gamma}(0.5,0.5),\quad
\boldsymbol{\Omega} \sim \mathcal{W}_J(\nu,\,\mathbf{T}),
$$

with $\nu=J+1$ and $\mathbf{T}=\mathrm{diag}(t^2)$.

We report posterior correlations at two levels:

* **population:** from $\mathbf{\Sigma}=\boldsymbol{\Omega}^{-1}$
* **individual:** from ${\boldsymbol{\theta}\_i}$ within each draw

```{r iw_model, message=FALSE, warning=FALSE}
modIWJags = "
model{
  for (i in 1:I){
    theta[i,1:J] ~ dmnorm(mu[], pSig2[,])
  }
  for (j in 1:J){
    mu[j]   ~ dnorm(mu_m, pow(mu_s, -2))
    pTau2[j] ~ dgamma(0.5, 0.5)
  }
  pSig2 ~ dwish(diagJ * pow(tuneT, 2), J + 1)

  for (n in 1:N){
    center[n] <- theta[sub[n], task[n]]
    y[n] ~ dnorm(center[n], pTau2[task[n]])
  }
}
"

runJags = function(mod, dat, pars, nchains = 2, niter = 3000, nburnin = 1000){
  jags(model.file = textConnection(mod),
       data = dat, n.chains = nchains, n.iter = niter,
       n.burnin = nburnin, n.thin = 1, parameters.to.save = pars)
}

iw_data = list(
  y     = dat$yc,
  task  = dat$JV,
  sub   = dat$S,
  I     = I,
  J     = J,
  N     = N,
  diagJ = diag(J),
  tuneT = 1,
  mu_m  = 0,
  mu_s  = 2
)

fitIW = runJags(modIWJags, iw_data, c("theta","pSig2"))
drawsIW = fitIW$BUGSoutput$sims.list
thetaIW = drawsIW$theta                 # M x I x J
pSig2IW = drawsIW$pSig2                 # M x J x J
M_IW = dim(thetaIW)[1]
```

**Correlations and intervals.**

```{r iw_corrs}
Ybar = tapply(dat$yc, list(dat$S, dat$JV), mean)
colnames(Ybar) = task_names
samp_cor = cor(Ybar); dimnames(samp_cor) = list(task_names, task_names)

pop_list = vector("list", M_IW)
for (m in 1:M_IW){
  Sigma = solve(pSig2IW[m,,])
  pop_list[[m]] = cov2cor(Sigma)
}

ind_list = vector("list", M_IW)
for (m in 1:M_IW){
  Th = thetaIW[m,,]
  colnames(Th) = task_names
  ind_list[[m]] = cor(Th)
}

summarize_corrs = function(L){
  A = abind::abind(L, along = 3)
  mean = apply(A, c(1,2), mean)
  lo   = apply(A, c(1,2), quantile, probs = .025)
  hi   = apply(A, c(1,2), quantile, probs = .975)
  dimnames(mean) = dimnames(lo) = dimnames(hi) = dimnames(samp_cor)
  list(mean = mean, lo = lo, hi = hi)
}

pop_sum = summarize_corrs(pop_list)
ind_sum = summarize_corrs(ind_list)

pairs_df = {
  idx = which(upper.tri(samp_cor), arr.ind = TRUE)
  data.frame(
    pair = paste0(rownames(samp_cor)[idx[,1]], "–", colnames(samp_cor)[idx[,2]]),
    samp = samp_cor[idx],
    pop_mean = pop_sum$mean[idx],
    pop_lo   = pop_sum$lo[idx],
    pop_hi   = pop_sum$hi[idx],
    ind_mean = ind_sum$mean[idx],
    ind_lo   = ind_sum$lo[idx],
    ind_hi   = ind_sum$hi[idx]
  )
}

# knitr::kable(
#   pairs_df,
#   align = "c",
#   caption = 
#     "Off-diagonal correlations: sample vs model (posterior means and 95% CIs).",
#   digits = 3
# )

head(pairs_df)
```

**Plots.** Heatmaps and interval comparison.

```{r iw_plots, fig.height=4.2, fig.width=11, message=FALSE, warning=FALSE}
cols = colorRampPalette(RColorBrewer::brewer.pal(11, "RdBu"))(201)
require(corrplot)
par(mfrow = c(1,3), mar = c(1.5,1.5,2,1))
corrplot(samp_cor,       
         method = "color", col = cols, cl.lim = c(-1,1),
         tl.col = "black", tl.cex = .7, addgrid.col = NA, title = "Sample", 
         mar = c(0,0,1,0))
corrplot(pop_sum$mean,   
         method = "color", col = cols, cl.lim = c(-1,1),
         tl.col = "black", tl.cex = .7, addgrid.col = NA, 
         title = "Model (population)", mar = c(0,0,1,0))
corrplot(ind_sum$mean,   
         method = "color", col = cols, cl.lim = c(-1,1),
         tl.col = "black", tl.cex = .7, addgrid.col = NA, 
         title = "Model (individual)",   mar = c(0,0,1,0))
```

```{r iw_forest, fig.height=9, fig.width=7}
ord = order(pairs_df$pop_mean, decreasing = TRUE)
y = seq_along(ord)
op = par(mar = c(4, 8.5, 2, 1))
plot(NA, xlim = c(-1, 1), ylim = c(0.5, length(y) + 0.5),
     xlab = "correlation", ylab = "", yaxt = "n")
abline(v = 0, lty = 2, lwd = 1)
segments(pairs_df$pop_lo[ord], y + .15, pairs_df$pop_hi[ord], y + .15, lwd = 3)
segments(pairs_df$ind_lo[ord], y - .15, 
         pairs_df$ind_hi[ord], y - .15, lwd = 3, col = "gray40")
points(pairs_df$pop_mean[ord], y + .15, pch = 19)
points(pairs_df$ind_mean[ord], y - .15, pch = 19, col = "gray40")
points(pairs_df$samp[ord], y, pch = 4, cex = 1, lwd = 2, col = "indianred2")
axis(2, at = y, labels = pairs_df$pair[ord], las = 2, cex.axis = 0.7)
legend("bottomleft",
       legend = c("population (mean ± CI)","individual (mean ± CI)","sample"),
       pch = c(19,19,4), col = c("black","gray40","indianred2"),
       pt.cex = c(1,1,0.9), bty = "n")
par(op)
```


---

# Hierarchical Bayesian factor model

Measurement model for indicator $j$ and person $i$:

$$
\begin{aligned}
y_{ijk}\mid \theta_{ij},\,\tau_j^2
&\sim \mathcal{N}\!\big(\theta_{ij},\,\tau_j^2\big),\\
\theta_{ij}\mid \mu_j,\,\boldsymbol{\lambda}_j,\,\delta_j^2,\,\boldsymbol{\eta}_i
&\sim \mathcal{N}\!\big(\mu_j+\boldsymbol{\lambda}_j^{\prime}\boldsymbol{\eta}_i,\,\delta_j^2\big),\\
\boldsymbol{\eta}_i &\sim \mathcal{N}_D(\mathbf{0},\mathbf{I}_D).
\end{aligned}
$$

Priors are weak normals on loadings and means, gammas on precisions. Identification uses constraint matrices: five task factors load on their two versions, two general factors are free.

```{r}
mod_bhfm = "
model {
  for (j in 1:J) {
    pTau2[j] ~ dgamma(0.5, 0.5)
    mu[j]    ~ dnorm(mu.m, pow(mu.s, -2))
    pDel2[j] ~ dgamma(0.5, 0.5 * pow(tuneDelta, 2))
    del2[j]  <- 1 / pDel2[j]

    for (d in 1:D) {
      lambda_pos[j,d]  ~ dnorm(0, pow(tuneLambda, -2)) T(0,)
      lambda_free[j,d] ~ dnorm(0, pow(tuneLambda, -2))
      aux_pos[j,d] <- constraint_zero[j,d] - constraint_pos[j,d]
      lambda[j,d] <- constraint_pos[j,d]*lambda_pos[j,d] + aux_pos[j,d]*lambda_free[j,d]
    }
  }

  for (i in 1:I) {
    for (d in 1:D) { eta[i,d] ~ dnorm(0, 1) }
  }

  for (i in 1:I) {
    for (j in 1:J) {
      center_theta[i,j] <- mu[j] + inprod(lambda[j,1:D], eta[i,1:D])
      theta[i,j] ~ dnorm(center_theta[i,j], pDel2[j])
    }
  }

  for (n in 1:N) {
    center[n] <- theta[sub[n], task[n]]
    y[n] ~ dnorm(center[n], pTau2[task[n]])
  }
}
"
```

**Constraints.**

```{r}
constraint_matrix_zero = matrix(0, nrow = 10, ncol = 7)
for (i2 in 1:5) constraint_matrix_zero[(i2*2-1):(i2*2), i2] = 1
constraint_matrix_zero[, 6:7] = 1

constraint_matrix_pos = matrix(0, nrow = 10, ncol = 7)
for (i2 in 1:5) constraint_matrix_pos[(i2*2-1), i2] = 1

constraint_matrix_zero; constraint_matrix_pos
```

**Fit.**

```{r}
data_list = list(
  y = dat$yc, task = dat$JV, sub = dat$S,
  I = I, J = J, N = N, D = D,
  constraint_zero = constraint_matrix_zero,
  constraint_pos  = constraint_matrix_pos,
  mu.m = 2, mu.s = 2, tuneDelta = 1, tuneLambda = 1
)
parameters = c("lambda","theta","mu","del2","pTau2","eta")
fit = runJags(mod_bhfm, data_list, parameters)
invisible(fit)
```

# Diagnostics (selected)

```{r fig.height=7, fig.width=8, message=FALSE, warning=FALSE}
sel = c("lambda[2,1]","lambda[4,2]","lambda[6,3]","del2[1]","pTau2[1]","mu[1]")
summ = fit$BUGSoutput$summary
keep = summ[sel, c("mean","sd","2.5%","50%","97.5%","Rhat","n.eff")]
knitr::kable(round(keep, 3), align = "c",
             caption = "Selected parameters: posterior summaries and diagnostics")

op = par(mfrow = c(3, 2), mar = c(3.2, 3.2, 2, 1))
for (v in sel) try(R2jags::traceplot(fit, varname = v), silent = TRUE)
par(op)
```

# Post-processing

```{r}
samples = fit$BUGSoutput$sims.list
lambda = samples$lambda
del2   = samples$del2

makePositive = function(lambda){
  Dloc = dim(lambda)[3]
  dimMean = apply(lambda, 3, mean)
  for (d in 1:Dloc){
    s = ifelse(dimMean[d] == 0, 1, sign(dimMean[d]))
    lambda[,,d] = s * lambda[,,d]
  }
  lambda
}

align = function(lambda, I = 0){
  M = dim(lambda)[1]; numFactors = dim(lambda)[3]
  eta = array(0, dim = c(M, I, numFactors))
  aligned = jointRot(lambda = lapply(1:M, function(x) lambda[x,,]),
                     eta    = lapply(1:M, function(x) eta[x,,]))
  aperm(abind(aligned$lambda, along = 3), c(3,1,2))
}

f2Eigenmax = function(lambda){
  euler2 = function(theta)
    matrix(c(cos(theta), -sin(theta), sin(theta), cos(theta)), 2, 2, byrow = TRUE)
  obj2 = function(theta, b){
    r = b %*% euler2(theta); apply(r^2, 2, sum)[1]
  }
  M = dim(lambda)[1]; pLam = apply(lambda, 2:3, mean)
  out = lambda
  ang = optimize(obj2, c(-pi, pi), maximum = TRUE, b = pLam)$maximum
  for (m in 1:M) out[m,,] = lambda[m,,] %*% euler2(ang)
  out
}

standardize_lambda = function(lambda, del2){
  M = dim(lambda)[1]; Jloc = dim(lambda)[2]
  lam_std = lambda; del2_std = del2
  Sigma = array(NA_real_, dim = c(M, Jloc, Jloc))
  Rho   = array(NA_real_, dim = c(M, Jloc, Jloc))
  for (m in 1:M){
    S = tcrossprod(lambda[m,,]) + diag(del2[m,], Jloc)
    Sigma[m,,] = S; Rho[m,,] = cov2cor(S)
    sigma = sqrt(diag(S))
    lam_std[m,,] = sweep(lambda[m,,], 1, sigma, "/")
    del2_std[m,] = del2[m,] / (sigma^2)
  }
  list(lambda = lam_std, del2 = del2_std, Sigma = Sigma, rho = Rho)
}
```

Align, standardize, summarize.

```{r}
lamb_g = lambda[,,6:7]
lamb_aligned = align(lamb_g, I = I)
lamb_aligned = f2Eigenmax(lamb_aligned)
lamb_aligned = makePositive(lamb_aligned)
lambda[,,6:7] = lamb_aligned

std = standardize_lambda(lambda, del2)
lam_std = std$lambda; del2_std = std$del2

makeTable = function(lambda, del2){
  M = dim(lambda)[1]; Jloc = dim(lambda)[2]; Dloc = dim(lambda)[3]
  lam_sum = matrix(0, Jloc, Dloc)
  uni_prop_task_mat = matrix(0, M, Jloc)
  fac_prop_total_mat = matrix(0, M, Dloc)
  uni_prop_total_vec = numeric(M)
  for (m in 1:M){
    tlam = lambda[m,,]; tdel = del2[m,]
    tot_com_task = rowSums(tlam^2)
    tot_task_j   = tot_com_task + tdel
    tot_com_fac  = colSums(tlam^2)
    tot_uni      = sum(tdel)
    tot_var      = sum(tdel + tot_com_task)
    lam_sum = lam_sum + tlam
    uni_prop_task_mat[m,] = tdel / tot_task_j
    fac_prop_total_mat[m,] = tot_com_fac / tot_var
    uni_prop_total_vec[m]  = tot_uni / tot_var
  }
  lam_mean = round(lam_sum / M, 3)
  uni_prop_task = round(colMeans(uni_prop_task_mat), 3)
  fac_prop_total = round(colMeans(fac_prop_total_mat), 3)
  uni_prop_total = round(mean(uni_prop_total_vec), 3)
  out = cbind(lam_mean, uni_prop_task)
  out = rbind(out, c(fac_prop_total, uni_prop_total))
  rownames(out) = c(task_names, "Prop. Var.")
  colnames(out) = c(paste0("F", 1:D), "Unique Var.")
  out
}

tab = makeTable(lam_std, del2_std)
knitr::kable(
  tab, align = "c", 
  caption = "Posterior mean loadings and variance proportions (standardized)."
)
```

# Plots

```{r fig.height=4.2, fig.width=11}
dims = dim(lam_std); M = dims[1]; Jloc = dims[2]; Dloc = dims[3]
fac_prop_total_mat = matrix(0, M, Dloc); uni_prop_total_vec = numeric(M)
for (m in 1:M){
  tlam = lam_std[m,,]; tdel = del2_std[m,]
  fac_prop_total_mat[m,] = colSums(tlam^2) / sum(tdel + rowSums(tlam^2))
  uni_prop_total_vec[m]  = sum(tdel) / sum(tdel + rowSums(tlam^2))
}
lam_mean = round(apply(lam_std, c(2,3), mean), 3)
base_colors = RColorBrewer::brewer.pal(5, "Set1")
cols = unlist(lapply(base_colors, function(col) c(adjustcolor(col, 0.6), col)))
task_names2 = c("Brentano","Ebbinghaus","Poggendorf","Ponzo","Zoellner")

par(mfrow = c(1,3), mar = c(5.2, 2.0, 2.5, 0.5))

# (A) mean loadings on general factors (F6,F7)
plot(NA, xlim = c(-1, 1), ylim = c(-1, 1), axes = FALSE, xlab = "", ylab = "")
grid(lwd = 2); box(); abline(h = 0, v = 0, lwd = 2)
for (i2 in 1:nrow(lam_mean)) segments(0, 0, lam_mean[i2, 6], 
                                      lam_mean[i2, 7], col = cols[i2], lwd = 2)
points(lam_mean[, 6], lam_mean[, 7], col = cols, pch = 19, cex = 1.1, lwd = 2)
legend("bottomleft", legend = task_names, col = cols, pch = 19, 
       cex = 0.75, bg = "white", ncol = 2)
mtext("(A)", side = 3, line = 0.4, cex = 0.85, font = 2, adj = 0)
mtext("General Factor 1", side = 1, line = 1.0, cex = 0.8)
mtext("General Factor 2", side = 2, line = 1.0, cex = 0.8)

# (B) variance share per task factor
plot(NA, xlim = c(0, 1), ylim = c(0, 50), axes = FALSE, xlab = "", ylab = "")
axis(1); box()
tcols = cols[c(1,3,5,7,9)]
for (j2 in 1:5){
  dens = density(fac_prop_total_mat[, j2])
  polygon(dens, col = adjustcolor(tcols[j2], 0.4), border = tcols[j2], lwd = 2)
}
legend("topright", legend = task_names2, fill = adjustcolor(tcols, 0.4),
       border = tcols, cex = 0.95)
mtext("(B)", side = 3, line = 0.4, cex = 0.85, font = 2, adj = 0)
mtext("Proportion of Total Variance (Task Factors)", 
      side = 1, line = 2.6, cex = 0.8)
mtext("Posterior Density", side = 2, line = 1.0, cex = 0.8)

# (C) general factors + unique variance
plot(NA, xlim = c(0, 1), ylim = c(0, 50), axes = FALSE, xlab = "", ylab = "")
axis(1); box()
tcols2 = c("black","gray60","brown")
for (d2 in 6:7){
  dens = density(fac_prop_total_mat[, d2])
  polygon(dens, 
          col = adjustcolor(tcols2[d2-5], 0.4), border = tcols2[d2-5], lwd = 2)
}
dens = density(uni_prop_total_vec)
polygon(dens, col = adjustcolor(tcols2[3], 0.35), border = tcols2[3], lwd = 2)
legend("topright", legend = c("General Factor 1",
                              "General Factor 2","Unique Variance"),
       fill = adjustcolor(tcols2[1:3], 0.35), border = tcols2[1:3], cex = 0.95)
mtext("(C)", side = 3, line = 0.4, cex = 0.85, font = 2, adj = 0)
mtext("Proportion of Total Variance", side = 1, line = 2.6, cex = 0.8)
mtext("Posterior Density", side = 2, line = 1.0, cex = 0.8)
```

**End.** This file provides the IW baseline and the factor model used in the paper.